---
output: github_document
---

<!-- examples.md is generated from examples.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(root.dir = "exec")
```

# rush examples

A comprehensive collection of examples showing what `rush` can do. Each section
demonstrates a specific argument or capability with real data and realistic
use cases.

```{r, include=FALSE}
# Set up example data for later sections.
system("rush run 'head(mtcars, 20)' > mtcars.csv 2>/dev/null")

system("rush run -d '\t' 'data.frame(city = c(\"Amsterdam\", \"Rotterdam\", \"Utrecht\", \"Den Haag\"), pop = c(921402, 655468, 361924, 552995), province = c(\"Noord-Holland\", \"Zuid-Holland\", \"Utrecht\", \"Zuid-Holland\"))' > cities.tsv 2>/dev/null")

writeLines(c("10", "20", "30", "40", "50"), "numbers.txt")

system("rush run --output flights.parquet 'data.frame(carrier = c(\"UA\",\"AA\",\"DL\",\"UA\",\"AA\",\"DL\",\"UA\",\"AA\",\"DL\",\"WN\"), origin = c(\"EWR\",\"JFK\",\"LGA\",\"EWR\",\"JFK\",\"LGA\",\"EWR\",\"JFK\",\"LGA\",\"EWR\"), dest = c(\"LAX\",\"SFO\",\"ATL\",\"ORD\",\"LAX\",\"DTW\",\"SFO\",\"MIA\",\"ATL\",\"DEN\"), dep_delay = c(-2, 5, 12, -1, 120, 3, 0, -4, 22, 45), arr_delay = c(-10, -2, 8, 5, 115, -5, -8, -12, 18, 50), air_time = c(320, 350, 120, 130, 330, 100, 340, 160, 115, 220), distance = c(2454, 2586, 762, 719, 2475, 502, 2565, 1089, 762, 1605))' 2>/dev/null")

system("rush run -l duckdb,DBI '{con <- DBI::dbConnect(duckdb::duckdb(), \"inventory.duckdb\"); DBI::dbWriteTable(con, \"products\", data.frame(id = 1:6, name = c(\"Widget\", \"Gadget\", \"Doohickey\", \"Thingamajig\", \"Whatsit\", \"Gizmo\"), category = c(\"A\", \"B\", \"A\", \"B\", \"A\", \"B\"), price = c(9.99, 24.50, 4.75, 18.00, 12.30, 31.00))); DBI::dbWriteTable(con, \"sales\", data.frame(product_id = c(1,1,2,3,4,4,5,6,6,6,1,2,3,5), quantity = c(5, 3, 2, 10, 1, 4, 7, 2, 1, 3, 8, 1, 6, 2), date = as.Date(\"2024-01-01\") + c(0,5,3,1,7,12,2,8,15,20,25,18,22,28))); DBI::dbDisconnect(con, shutdown = TRUE); invisible()}' 2>/dev/null")
```

---

## rush run

### Simple expressions

At its core, `rush run` evaluates any R expression and prints the result.

```{bash}
rush run '6 * 7'
```

Multiple semicolon-separated expressions work; the value of the last one is printed:

```{bash}
rush run 'x <- 1:10; mean(x)'
```

Strings, vectors, and lists all print naturally:

```{bash}
rush run 'paste("Hello", "from", "R", sep = ", ")'
```

```{bash}
rush run 'rev(LETTERS[1:8])'
```

### Reading a single file

Pass a file path after the expression, and it's read into a data frame called `df`:

```{bash}
rush run 'nrow(df)' mtcars.csv
```

With just a file and no expression, the data is simply printed — a quick way to peek:

```{bash}
rush run 'head(df, 5)' mtcars.csv
```

### Reading from standard input (`-`)

Use `-` to read from stdin. This makes `rush` composable in shell pipelines:

```{bash}
echo "name,score
Alice,92
Bob,85
Carol,97" | rush run 'df |> dplyr::arrange(dplyr::desc(score))' -
```

You can pipe the output of one `rush` into another:

```{bash}
rush run 'head(mtcars, 5)' | rush run 'dplyr::select(df, mpg, hp, wt)' -
```

### Reading multiple files

When you pass multiple files, each is read into a list called `dfs`, keyed by the file's base name:

```{bash}
rush run 'names(dfs)' mtcars.csv cities.tsv
```

```{bash}
rush run 'nrow(dfs[["mtcars"]])' mtcars.csv cities.tsv
```

### `--delimiter` (`-d`)

Override the delimiter for both reading and writing. Here we read a tab-separated file:

```{bash}
rush run -d "\t" 'head(df, 3)' cities.tsv
```

And write tab-separated output:

```{bash}
rush run -d "\t" 'head(mtcars, 3)' | cat
```

### `--no-header` (`-H`)

Read a file that has no header row. Columns are auto-named `x1`, `x2`, ...:

```{bash}
rush run -H 'sum(df$x1)' numbers.txt
```

```{bash}
seq 5 | rush run -H 'cumsum(df$x1)' -
```

### `--no-clean-names` (`-C`)

By default, `rush` passes data frames through `janitor::clean_names()` to
normalize column names (lowercase, underscores). Disable this with `-C` to
preserve the original names exactly:

```{bash}
echo "First Name,Last Name,Score (%)
Alice,Smith,92" | rush run 'names(df)' -
```

```{bash}
echo "First Name,Last Name,Score (%)
Alice,Smith,92" | rush run -C 'names(df)' -
```

### `--tidyverse` (`-t`)

Load the entire Tidyverse (plus glue) so you can use dplyr, tidyr, stringr, etc.
without qualifying names:

```{bash}
rush run -t 'mtcars |> group_by(cyl) |> summarise(mean_mpg = mean(mpg)) |> arrange(cyl)' mtcars.csv
```

### `--library` (`-l`)

Load specific packages. Comma-separate multiple packages:

```{bash}
rush run -l stringr 'stringr::str_to_title("hello world")'
```

```{bash}
rush run -l dplyr,tidyr 'df |>
  tidyr::pivot_longer(cols = c(mpg, hp), names_to = "metric") |>
  dplyr::slice_head(n = 4)' mtcars.csv
```

### `--seed`

Set the random seed for reproducible results:

```{bash}
rush run --seed 42 'sample(1:100, 5)'
```

Run it again — same seed, same result:

```{bash}
rush run --seed 42 'sample(1:100, 5)'
```

### `--output` (`-o`)

Write the result to a file instead of stdout. The output format is inferred from
the extension:

```{bash}
rush run -o summary.csv 'data.frame(stat = c("min", "mean", "max"),
                                     mpg = c(min(df$mpg), mean(df$mpg), max(df$mpg)))' mtcars.csv
cat summary.csv
```

Write to Parquet:

```{bash}
rush run -o subset.parquet 'dplyr::filter(df, mpg > 25)' mtcars.csv
rush run 'df' subset.parquet
```

### `--dry-run` (`-n`)

Print the generated script instead of running it. This is great for learning
what `rush` does under the hood, or for saving the script to a file:

```{bash}
rush run -n 'df |> dplyr::filter(mpg > 25)' mtcars.csv
```

### `--no-ir` (`-I`)

Run the generated script with plain `Rscript` instead of `ir`. No package
resolution happens — everything must already be installed:

```{bash}
rush run -I '2 + 2'
```

### `--verbose` (`-v`)

Show the parsed arguments before running. Useful for debugging complex
invocations:

```{bash}
rush run -v --seed 7 'sum(1:10)' 2>&1
```

### Reading Parquet files

Files ending in `.parquet` or `.pq` are read with nanoparquet automatically:

```{bash}
rush run 'df |> dplyr::filter(dep_delay > 10) |> dplyr::select(carrier, origin, dest, dep_delay)' flights.parquet
```

### Reading DuckDB databases

Point `rush` at a `.duckdb` file and all tables are loaded into `dfs`:

```{bash}
rush run 'lapply(dfs, head, 3)' inventory.duckdb
```

---

## rush sql

### Basic queries

Run a SQL query against one or more files. Each file becomes a DuckDB relation:

```{bash}
rush sql "SELECT mpg, cyl, hp FROM mtcars WHERE mpg > 25 ORDER BY mpg DESC" mtcars.csv
```

### Aggregations

DuckDB's full SQL is available — aggregations, window functions, CTEs:

```{bash}
rush sql "SELECT cyl,
                 COUNT(*) AS n,
                 ROUND(AVG(mpg), 1) AS avg_mpg,
                 MIN(hp) AS min_hp,
                 MAX(hp) AS max_hp
          FROM mtcars
          GROUP BY cyl
          ORDER BY cyl" mtcars.csv
```

### Querying Parquet files

Parquet files are registered via DuckDB's `read_parquet`, so large-file queries
run without loading everything into memory:

```{bash}
rush sql "SELECT carrier, COUNT(*) AS flights, ROUND(AVG(arr_delay), 1) AS avg_delay
          FROM flights
          GROUP BY carrier
          ORDER BY avg_delay DESC" flights.parquet
```

### Querying DuckDB databases

A `.duckdb` file is attached as a schema, and its tables are accessible as
`schema.table`:

```{bash}
rush sql "SELECT p.name, p.category, SUM(s.quantity) AS total_sold
          FROM inventory.products p
          JOIN inventory.sales s ON p.id = s.product_id
          GROUP BY p.name, p.category
          ORDER BY total_sold DESC" inventory.duckdb
```

### Querying standard input

Use `-` to read CSV from stdin as a relation named `stdin`:

```{bash}
echo "item,price,qty
apple,1.20,10
banana,0.80,25
cherry,3.50,5" | rush sql "SELECT item, price * qty AS total FROM stdin ORDER BY total DESC" -
```

### Joining multiple files

Pass multiple files and each becomes a relation you can join:

```{bash}
echo "cyl,fuel_type
4,regular
6,premium
8,premium" > fuel.csv
rush sql "SELECT m.cyl, f.fuel_type, ROUND(AVG(m.mpg), 1) AS avg_mpg
          FROM mtcars m JOIN fuel f ON m.cyl = f.cyl
          GROUP BY m.cyl, f.fuel_type
          ORDER BY m.cyl" mtcars.csv fuel.csv
```

### SQL with `--output`

Pipe a SQL result to Parquet or another file:

```{bash}
rush sql -o efficient.parquet "SELECT * FROM mtcars WHERE mpg > 20" mtcars.csv
rush run 'nrow(df)' efficient.parquet
```

### SQL with `--tidyverse`

Load tidyverse packages alongside the SQL query — useful when you want to
post-process the result with `--library` or pipe it into further R:

```{bash}
rush sql -t "SELECT cyl, AVG(mpg) as avg_mpg FROM mtcars GROUP BY cyl" mtcars.csv
```

### SQL `--dry-run`

See the generated script that runs the query:

```{bash}
rush sql -n "SELECT COUNT(*) FROM mtcars" mtcars.csv
```

---

## rush plot

### Scatter plot (default with x and y)

When both `--x` and `--y` are given, `rush` guesses `geom_point`:

```{bash}
rush plot -n -x wt -y mpg mtcars.csv
```

### Histogram (default with x only)

When only `--x` is given, `rush` guesses `geom_histogram`:

```{bash}
rush plot -n -x mpg mtcars.csv
```

### `--geom` (`-g`)

Override the guessed geom. Any ggplot2 geom suffix works:

```{bash}
rush plot -n -x mpg -g density mtcars.csv
```

```{bash}
rush plot -n -x cyl -y mpg -g boxplot mtcars.csv
```

```{bash}
rush plot -n -x wt -y mpg -g smooth mtcars.csv
```

### `--color` (`-c`)

Map a column to the color aesthetic:

```{bash}
rush plot -n -x wt -y mpg -c 'factor(cyl)' mtcars.csv
```

### `--fill` (`-f`)

Map a column to fill (useful for histograms, bar charts, density plots):

```{bash}
rush plot -n -x mpg -g density -f 'factor(cyl)' --alpha 0.5 mtcars.csv
```

### `--alpha` (`-a`)

Map a column to transparency, or use a constant for semi-transparent geoms:

```{bash}
rush plot -n -x wt -y mpg --size hp --alpha 0.6 mtcars.csv
```

### `--size`

Map a column to point size — great for bubble charts:

```{bash}
rush plot -n -x wt -y mpg --size hp -c 'factor(cyl)' mtcars.csv
```

### `--shape`

Map a column to point shape:

```{bash}
rush plot -n -x wt -y mpg --shape 'factor(vs)' mtcars.csv
```

### `--group`

Group observations for geoms like `geom_line` that connect points:

```{bash}
rush plot -n -x wt -y mpg --group 'factor(cyl)' -g line mtcars.csv
```

### `--aes`

Pass additional aesthetics as `key=value` pairs for anything not covered by
the dedicated flags:

```{bash}
rush plot -n -x wt -y mpg --aes "label=rownames(df)" -g text mtcars.csv
```

### `--log`

Log-transform axes. Accepts `x`, `y`, or `xy`:

```{bash}
rush plot -n -x hp -y mpg --log xy mtcars.csv
```

```{bash}
rush plot -n -x hp -y mpg --log x mtcars.csv
```

### `--facets` (one-sided: facet_wrap)

A one-sided formula uses `facet_wrap`:

```{bash}
rush plot -n -x wt -y mpg --facets '~ cyl' mtcars.csv
```

### `--facets` (two-sided: facet_grid) + `--margins`

A two-sided formula uses `facet_grid`. Add `--margins` to show marginal panels:

```{bash}
rush plot -n -x wt -y mpg --facets 'vs ~ cyl' --margins mtcars.csv
```

### `--title`, `--xlab`, `--ylab`

Add axis labels and a title:

```{bash}
rush plot -n -x wt -y mpg --title "Fuel Efficiency vs Weight" --xlab "Weight (1000 lbs)" --ylab "Miles per Gallon" mtcars.csv
```

### `--pre`

Run code before the plot is built. Useful for data transformations:

```{bash}
rush plot -n -x cyl -y mean_mpg --pre 'df <- df |> dplyr::summarise(mean_mpg = mean(mpg), .by = cyl)' -g col mtcars.csv
```

### `--post`

Run code after the plot is built. The plot object is available as `p`, and
the value of the last `--post` expression becomes the result. Use this to
add custom layers, themes, or annotations:

```{bash}
rush plot -n -x wt -y mpg --post 'p + ggplot2::theme_minimal() + ggplot2::geom_smooth(method = "lm")' mtcars.csv
```

### Saving to a file (`--output`)

When `--output` points to a file, the plot is saved (device inferred from extension):

```{bash}
rush plot -x wt -y mpg -o scatter.png mtcars.csv 2>/dev/null
ls -la scatter.png
```

### `--width`, `--height`, `--units`, `--dpi`

Control the saved image dimensions and resolution:

```{bash}
rush plot -x wt -y mpg -o hires.png --width 10 --height 6 --units in --dpi 150 mtcars.csv 2>/dev/null
ls -la hires.png
```

### Reading from stdin for plots

Without a file argument, `plot` reads from standard input:

```{bash}
rush run 'head(mtcars, 10)' | rush plot -n -x wt -y mpg -
```

### Plotting multiple files

With multiple files, each is read into `dfs`. Use `--pre` to combine them
into `df` before plotting:

```{bash}
echo "src,x,y
A,1,2
A,2,4
A,3,5" > a.csv
echo "src,x,y
B,1,3
B,2,2
B,3,6" > b.csv
rush plot -n -x x -y y -c src --pre 'df <- dplyr::bind_rows(dfs)' a.csv b.csv
```

### Plotting Parquet data

Parquet files work seamlessly:

```{bash}
rush plot -n -x distance -y air_time -c carrier flights.parquet
```

### `--tidyverse` with plot

Load the tidyverse for richer transformations in `--pre` or `--post`:

```{bash}
rush plot -n -t -x cyl -y n --pre 'df <- count(df, cyl)' -g col mtcars.csv
```

### `--seed` with plot

Set the random seed when your plot involves sampling or jitter:

```{bash}
rush plot -n --seed 123 -x cyl -y mpg -g jitter mtcars.csv
```

---

## Combining commands in pipelines

The real power of `rush` is combining commands in Unix pipelines:

```{bash}
# Query with SQL, then plot the result
rush sql "SELECT cyl, COUNT(*) as n, AVG(mpg) as avg_mpg FROM mtcars GROUP BY cyl" mtcars.csv | \
  rush plot -n -x cyl -y avg_mpg --size n -g point -
```

```{bash}
# Filter with rush run, convert to Parquet, query the Parquet
rush run 'dplyr::filter(df, mpg > 15)' mtcars.csv | \
  rush sql "SELECT cyl, ROUND(AVG(mpg),1) as avg FROM stdin GROUP BY cyl ORDER BY cyl" -
```

```{r, include=FALSE}
unlink(c("mtcars.csv", "cities.tsv", "numbers.txt", "flights.parquet",
         "inventory.duckdb", "fuel.csv", "a.csv", "b.csv", "summary.csv",
         "subset.parquet", "efficient.parquet", "scatter.png", "hires.png"))
```
