---
title: "Getting Started with rush"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rush}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rush)
```

## Overview

rush lets you run R expressions, query data with SQL, create plots, and convert
between file formats -- all from a single function call. Each function assembles
a self-contained R script and executes it with
[ir](https://r-lib.github.io/ir/), which resolves package dependencies on the
fly.

The four main functions are:

- `rush_run()` -- run an R expression on data files
- `rush_sql()` -- query files with SQL via DuckDB
- `rush_plot()` -- create a quick ggplot2 visualization
- `rush_convert()` -- convert between file formats

All examples below use `dry_run = TRUE` to show the generated script without
executing it (which would require `ir` to be installed).

## Running R expressions

The simplest use case is evaluating an R expression:

```{r}
rush_run("1 + 1", dry_run = TRUE)
```

When you provide input files, they're read into `df` (the first file) and `dfs`
(a named list of all files):
```{r}
rush_run("head(df)", file = "data.csv", dry_run = TRUE)
```

Multiple files are supported. The format is detected from the file extension:

```{r}
rush_run("nrow(dfs$sales)", file = c("sales.parquet", "products.csv"),
         dry_run = TRUE)
```

You can write results to a file -- the output format is inferred from the
extension:

```{r}
rush_run("dplyr::filter(df, amount > 100)",
         file = "sales.csv", output = "big_sales.parquet",
         dry_run = TRUE)
```

### Loading packages

Use `library` to load packages, or `tidyverse = TRUE` as a shortcut:

```{r}
rush_run("df |> filter(x > 0) |> mutate(y = x^2)",
         file = "data.csv", tidyverse = TRUE, dry_run = TRUE)
```

## Querying with SQL

`rush_sql()` runs a DuckDB SQL query. Input files are registered as relations
that you can reference in your query:

```{r}
rush_sql("SELECT name, SUM(amount) AS total FROM sales GROUP BY name",
         file = "sales.csv", dry_run = TRUE)
```

You can join across files of different formats:

```{r}
rush_sql("SELECT s.*, p.category FROM sales s JOIN products p USING (id)",
         file = c("sales.csv", "products.parquet"), dry_run = TRUE)
```

## Creating plots

`rush_plot()` generates a ggplot2 visualization. In a terminal it renders as
ANSI art; with an output file it saves as PNG, PDF, etc.:

```{r}
rush_plot("mtcars.csv", x = "wt", y = "mpg", dry_run = TRUE)
```

The geometry is guessed automatically (`geom_point` when both x and y are
given, `geom_histogram` for just x), but you can override it:

```{r}
rush_plot("mtcars.csv", x = "mpg", color = "cyl",
          geom = "density", dry_run = TRUE)
```

Faceting and labels:

```{r}
rush_plot("mtcars.csv", x = "wt", y = "mpg",
          facets = "~ cyl", title = "Cars by Cylinder Count",
          xlab = "Weight", ylab = "Miles per Gallon", dry_run = TRUE)
```

## Converting between formats

`rush_convert()` reads one or more files and writes them in a different format:

```{r}
rush_convert("data.csv", output = "data.parquet", dry_run = TRUE)
```

For multiple files, use a template with `%(file_name)s` or `%(file_index)d`:

```{r}
rush_convert(c("jan.csv", "feb.csv"),
             output = "%(file_name)s.parquet", dry_run = TRUE)
```

## Supported formats

rush supports a wide range of formats for both reading and writing:

| Format | Extensions | Read | Write |
|--------|-----------|------|-------|
| Delimited (CSV/TSV) | .csv, .tsv, .txt | Yes | Yes |
| Parquet | .parquet, .pq | Yes | Yes |
| JSON | .json | Yes | Yes |
| JSONL/NDJSON | .jsonl, .ndjson | Yes | Yes |
| Excel | .xlsx, .xls | Yes | Yes |
| Arrow IPC | .arrow, .feather, .ipc | Yes | Yes |
| DuckDB | .duckdb, .ddb | Yes | Yes |
| SQLite | .sqlite, .db | Yes | Yes |
| SPSS | .sav, .zsav, .por | Yes | Yes |
| Stata | .dta | Yes | Yes |
| SAS | .sas7bdat, .xpt | Yes | Yes |
| RDS | .rds | Yes | Yes |
| ODS | .ods | Yes | Yes |
| FASTA | .fasta, .fa, .fna | Yes | Yes |
| FASTQ | .fastq, .fq | Yes | Yes |
| YAML | .yaml, .yml | Yes | Yes |
| TOML | .toml | Yes | Yes |
| XML | .xml | Yes | Yes |
| Fixed-width | .fwf | Yes | -- |
